Machine Learning Models for the Segmentation of Clinical Exchange

### 2018-03-08 paper-amia-2018/manuscript.tex
update introduction

Rapidly developing computational technologies, specifically, machine learning models, offer a unique opportunity to accelerate this process. A few studies were done for the automatic annotation of clinical interactions in the setting of MI intervention. Lacson et al.\cite{lacson2005automatic} applied AdaBoost\cite{freund1999short} classifier for annotating the interactions in hemodialysis phone dialog into 3 categories. Our research group has recently applied several machine learning-based models to MI sessions \cite{hasan2016study,kotov2015interpretable}. A simple communication code scheme was automated to characterize patient communication and achieved accuracy comparable to human coders\cite{hasan2016study}.

Experimental data utilized in the above MI studies were prepared by transcribing audio conversation which involves a counselor, a patient, and some cases a caregiver\cite{hasan2016study,kotov2015interpretable}. However, this study utilized e-coaching data which have different structure and context compared to clinical data. E-coaches use email to deliver motivation-enhancing coaching to encourage healthy eating, grounded in the principles of motivational interviewing. E-coaching data is comprised of email responses which are unsegmented, unlike more traditional dyadic clinical communication where segmentation occurs naturally due to its conversation nature. Discourse analysis on e-coaching text can help to understand the communication process in e-coaching text by revealing the socio-psychological characteristics of a patient\cite{siegfried1995therapeutic}. By utilizing e-coaching data and leveraging innovative machine learning models, the ultimate goal of this research study is to fully automate the communication coding process.

A significant barrier to fully automated the communication coding process is the segmentation of the text data. During traditional qualitative coding, coders determine where to segment text, that is, where one code ends and another begins. Automatic segmentation of e-coaching intervention sessions is a challenging task due to the 2 important reasons. First, the email is an unstructured text that contains informal email exchange in non-traditional formats. Second, a text segment does not necessarily belong to the entire sentence or collection of sentences. One sentence can be segmented into multiple MI behaviors or several sentences may represent a single MI code. Figure~\ref{fig:text-segment} illustrates the segmentation of an e-coaching email exchange where the first sentence segmented into 2 different MI behaviors. On the other hand, fourth and fifth segments contain only one and multiple sentences, respectively.

@@

Recently, an online clinical intervention called MENU GenY\cite{alexander2017motivations} (Making Effective Nutrition Choices for Generation Y) was tested. MENU GenY is a technology-based public health intervention to encourage increased fruit and vegetable intake among young adult age 21-30, utilized personalized e-coaching data. The main focus of MENU GenY is to develop a better coding dictionary among GenY to improve eating habits. However, segmentation of clinical conversation in the context of electronically delivered interventions, in particular, segmentation of clinical interaction text into groups of MI behaviors, is still ignored while relying on manual hand-coded approach. Therefore, this study introduces a novel approach and the authors are not aware of any other work this approach has been considered for the segmentation of e-coaching text.

@@

For the experiment, we utilized three type of features including word (textual feature), topic, and punctuation mark. Each word represented in a binary format, where 1 indicates the appearance of the word and 0 for absence. Topics are considered as features since topic models are very effective\cite{kotov2015interpretable,hashimoto2016topic,lu2016modeling} to represent text documents. In this paper, we exploit a topic model named Labeled LDA\cite{ramage2009labeled} model, where topics are corresponds to labels. The generative process of Labeled LDA drawing the multinomial topic distributions over vocabulary for each topic. In our experiment, we represent each word in a vector of 2 topics, where the number of topics is experimentally determined by the model performance\cite{kotov2015interpretable}. Punctuation mark containing one of the symbols \{`.', `,', `!', `?', `:', `;', `-'\} is also employed as feature. This is one of the most important features as they indicate the boundary of a sentence, clause, and phrase.

@@

As follows from Table~\ref{tab:result_boundary}, KNN performs best among all machine learning models in terms of precision and F1-measure, achieved 0.808 precision with 0.728 F1-measure when contextual features are used, and 0.820 precision with 0.779 F1-measure when a combination of contextual, topic, and punctuation mark features are used. However, RNN demonstrates the lowest performance among all models in terms of recall and F1-measure while GRU shows 3.72\%, 17.79\% and 11.47\% higher precision, recall, and F1-measure than LSTM. In this study, SVM appears as the second highest model in terms of precision and F1-measure, obtains highest 0.679 recall when only textual features are used. On the other hand, NB exhibits lowest precision value 0.594, provides better recall and F1-measure than RNN model. When textual features are used in combination with topic and punctuation mark, recall increases by 0.6\%, 2.5\%, and 11.92\%; and F1-measure increases by 0\%, 3.39\%, and 7\% for NB, SVM, and KNN models, respectively. Nevertheless, precision increases by 4.31\% and 1.49\% for SVM and KNN methods while decreases by 0.7\% in NB. \\

@@

In this paper, results are reported by each class to avoid confusion about the overall model performance. In addition, standard metrics: precision, recall, and F1-measure were used to eliminate doubt about the model performance because accuracy is misleading for imbalance dataset. AUC values are also outlined due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. NB shows the highest AUC values, achieved 0.978 for both cases while provides lowest classification results except RNN. On the other hand, KNN and SVM exhibit 0.972 and 0.835 AUCs when only textual features are used; and 0.959 and 0.844 AUCs when a combination of textual, topic and punctuation mark features are used. Finally, RNN demonstrates lowest AUC values among all machine learning models, achieved AUC values 0.704 and 0.740 for LSTM and GRU, respectively. The conclusion drawn from the ROC curves, also confirmed the robustness and superiority of KNN model for the segmentation of clinical exchange.

We observed worst results of RNN, in particular, LSTM and GRU for the text segmentation. We believe that RNN performed poorly because it has a large set of weights which required longer sequence with a large set of data for all labels. In this study, maximum length of the input sequence was 3 and we utilized 3,138 examples of boundary case which failed to achieve good results. While the performance of RNN is poor, GRU performs better than LSTM which was observed in other previous study\cite{chung2014empirical}.

@@

The limitation of this study is that e-coaching text is collected from a single medical institute; formatting, style, and email segment can be different in other settings. Therefore, there is a need to replicate the experiments with different data sets. As our future work, we plan to evaluate our approach on other datasets involves in discourse analysis. We also plan to use more relevant features to improve model performance. For example, part-of-speech tagging\cite{hasan2016feedback} and distance from the begining and end of the current sentence, might significantly enhance the classification performance.

\section*{Conclusion}
Segmentation of e-coaching text is an integral part of developing an automated e-coaching intervention. Although several studies have done in clinical interventions, they are limited by the qualitative coding of clinical interactions. In addition, previous studies in medical domain, mainly segmented clinical text into sections and sentences, none of them are used for the segmentation of text into groups of MI behaviors in the setting of discourse analysis with email under the principle of motivational interviews. In this paper, we compared the performance of machine learning models for the task of segmentation of e-coaching text. We found out that k-nearest neighbor provides the best performance for the segmentation of text in terms of all performance metrics. Manual segmentation of e-coaching data is very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify individual text segments, which can be annotated directly with a classification model. This approach will also help for developing fully automated e-coaching and accelerate the pace of identifying effective communication strategies.

>>>>

Rapidly developing computational technologies, specifically, machine learning models, offer a unique opportunity to accelerate this process. Machine learning-based classification methods have been successfully applied to a variety of analytical tasks on textual data including classification\cite{nigam2000text}, sentiment analysis\cite{wang2012baselines}, and digital forensics\cite{de2002language}. A few studies have been done for the automatic annotation of clinical interactions in the setting of MI intervention. Lacson et al.\cite{lacson2005automatic} applied AdaBoost\cite{freund1999short} classifier for annotating the interactions in hemodialysis phone dialog as Clinical, Technical,
Backchannel, and Miscellaneous categories. Our research group has recently applied several machine learning-based models to MI sessions \cite{hasan2016study,kotov2015interpretable}. A simple communication code scheme was automated to characterize patient communication and achieved accuracy comparable to human coders\cite{hasan2016study}.

Experimental data utilized in the above MI studies were prepared by transcribing audio conversation which involves a counselor, a patient, and some cases a caregiver\cite{hasan2016study,kotov2015interpretable}. However, this study utilized e-coaching data which have different structure and context compared to traditional clinical data. E-coaches use email to deliver motivation-enhancing coaching to encourage healthy eating, grounded in the principles of motivational interviewing. E-coaching data is comprised of email responses which are unsegmented, unlike more traditional dyadic clinical communication where segmentation occurs naturally due to its conversation nature. Discourse analysis on e-coaching text can help to understand the communication process in e-coaching text by revealing the socio-psychological characteristics of a patient\cite{siegfried1995therapeutic,kalchbrenner2013recurrent,pierre2016neural}. By utilizing e-coaching data and leveraging innovative machine learning models, the ultimate goal of this research study is to fully automate the communication coding process.

A significant barrier to fully automated the communication coding process is the segmentation of the text data. During traditional qualitative coding, coders determine where to segment text, that is, where one code ends and another begins. Automatic segmentation of e-coaching intervention sessions is a challenging task due to the 2 important reasons. First, the email is an unstructured text that contains informal email exchange as a non-traditional format. Second, a text segment does not necessarily belong to the entire sentence or collection of sentences. One sentence can be segmented into multiple MI behaviors or several sentences may represent a single MI code. Figure~\ref{fig:text-segment} illustrates the segmentation of an e-coaching email exchange where the first sentence segmented into 2 different MI behaviors. On the other hand, fourth and fifth segments contain 1 and 3 sentences, respectively.

@@

Recently, an online clinical intervention called MENU GenY\cite{alexander2017motivations} (Making Effective Nutrition Choices for Generation Y) was proposed to test its efficacy. MENU GenY is a technology-based public health intervention to encourage increased fruit and vegetable intake among young adult age 21-30, utilized personalized e-coaching data. The goal of MENU GenY was to develop a better coding dictionary among GenY to improve eating habits. However, segmentation of clinical conversation in the context of electronically delivered interventions, in particular, segmentation of clinical interaction text into groups of MI behaviors, is still ignored while relying on manual hand-coded approach. Therefore, this study introduces a novel approach and the authors are not aware of any other work this approach has been considered for the segmentation of e-coaching text.

@@

For the experiment, we utilized three type of features including word (textual feature), topic, and punctuation mark. Each word represented in a binary format, where 1 indicates the appearance of the word and 0 for absence. Topics are considered as features since topic models are very effective\cite{kotov2015interpretable,hashimoto2016topic,lu2016modeling} to represent text documents. In this paper, we exploit a topic model named Labeled LDA\cite{ramage2009labeled} model, where topics correspond to labels. The generative process of Labeled LDA drawing the multinomial topic distributions over vocabulary for each topic. In our experiment, we represent each word in a vector of 2 topics, where the number of topics is experimentally determined by the model performance\cite{kotov2015interpretable}. Punctuation mark containing one of the symbols \{`.', `,', `!', `?', `:', `;', `-'\} is also employed as feature. This is one of the most important features as they indicate the boundary of a sentence, clause, and phrase.

@@

As follows from Table~\ref{tab:result_boundary}, KNN performs best among all machine learning models in terms of precision and F1-measure, achieved 0.808 precision with 0.728 F1-measure when contextual features are used, and 0.820 precision with 0.779 F1-measure when a combination of contextual, topic, and punctuation mark features are used. However, RNN demonstrates the lowest performance among all models in terms of recall and F1-measure while GRU shows 3.72\%, 17.79\% and 11.47\% higher precision, recall, and F1-measure than LSTM. In this study, SVM appears as the second highest model in terms of precision and F1-measure, obtains highest 0.679 recall when only textual features are used. On the other hand, NB exhibits lowest precision value 0.594, provides better recall and F1-measure than RNN model. When textual features are used in combination with topic and punctuation mark features, recall increases by 0.6\%, 2.5\%, and 11.92\%; and F1-measure increases by 0\%, 3.39\%, and 7\% for NB, SVM, and KNN models, respectively. Nevertheless, precision increases by 4.31\% and 1.49\% for SVM and KNN methods while decreases by 0.7\% in NB. \\

@@

In this paper, results are reported by each class to avoid confusion about the overall model performance. In addition, standard metrics: precision, recall, and F1-measure were used to eliminate doubt about the model performance because accuracy is misleading for imbalance dataset. AUC values are also outlined due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. NB shows the highest AUC values, achieved 0.978 for both cases while provides lowest classification results except RNN. On the other hand, KNN and SVM exhibit 0.972 and 0.835 AUCs when only textual features are used; and 0.959 and 0.844 AUCs when a combination of textual, topic and punctuation mark features are used. Finally, RNN demonstrates lowest AUC values among all machine learning models, achieved AUC values 0.704 and 0.740 for LSTM and GRU, respectively. The conclusion drawn from the ROC curves also confirmed the robustness and superiority of KNN model for the segmentation of clinical exchange.

We observed worst results of RNN, in particular, LSTM and GRU for the text segmentation. We believe that RNN performed poorly because it has a large set of weights which required longer sequence with a large set of data for all labels. In this study, a maximum length of the input sequence was 3 and we utilized 3,138 examples of boundary case which failed to achieve good results. While the performance of RNN is poor, GRU performs better than LSTM which was observed in other previous study\cite{chung2014empirical}.

@@

The limitation of this study is that e-coaching text is collected from a single medical institute; formatting, style, and email segment can be different in other settings. Therefore, there is a need to replicate the experiments with different data sets. As our future work, we plan to evaluate our approach on other datasets involves in discourse analysis. We also plan to use more relevant features to improve model performance. For example, part-of-speech tagging\cite{hasan2016feedback} and distance from the beginning and end of the current sentence might significantly enhance the classification performance.

\section*{Conclusion}
Segmentation of e-coaching text is an integral part of developing an automated e-coaching intervention. Although several studies have done in clinical interventions, they are limited by the qualitative coding of clinical interactions. In addition, previous studies in the medical domain mainly segmented clinical text into sections and sentences, none of them are used for the segmentation of text into groups of MI behaviors in the setting of discourse analysis with email under the principle of motivational interviews. In this paper, we compared the performance of machine learning models for the task of segmentation of e-coaching text. We found out that k-nearest neighbor provides the best performance for the segmentation of text in terms of all performance metrics. Manual segmentation of e-coaching data is very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify individual text segments, which can be annotated directly with a classification model. This approach will also help for developing fully automated e-coaching and accelerate the pace of identifying effective communication strategies.
